---
title: "malaria_analysis"
author: "Yixin"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```

Load required library
```{r message=FALSE,warning = FALSE }
library(plotly)
library(dplyr)
library(readr)
library(tidyverse)
library(countrycode)
```

Read the 3 datasets from Github

```{r message=FALSE, warning = FALSE}
malaria_death <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_deaths.csv", show_col_types = FALSE)

malaria_death_age <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_deaths_age.csv", show_col_types = FALSE)

malaria_inc <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018/2018-11-13/malaria_inc.csv", show_col_types = FALSE)

```

## Exploratory Analysis
First step : Look at malaria_death.csv data to have a quick overview of the situation across the world

```{r}
# basic data cleaning by rename the column and filter out NA values
malaria_death_processed <- malaria_death %>%
  setNames(c("country","code","year","deaths")) %>%
  filter(!is.na(code))
```

Summary can provide us with basic statistical measurements like mean, minimum and maximum value of malaria deaths

```{r}
summary(malaria_death_processed)
```

Randomly choose six countries to see malaria deaths trend over years 

```{r}
malaria_death_processed %>%
  filter(country %in% sample(unique(country),6)) %>%
  ggplot(aes(year, deaths, color = country)) +
  geom_line() +
  labs(x = "Year", y = "Deaths per 100,000")
```

Based on the chart above, we can improve by visualizing Top six countries with highest average deaths

```{r}
top_six_death_country <- malaria_death_processed %>% 
  group_by(country) %>%
  summarise(avg_deaths = mean(deaths)) %>%
  arrange(desc(avg_deaths)) %>%
  head()

malaria_death_processed %>%
  filter(country %in% top_six_death_country$country) %>%
  ggplot(aes(year, deaths, color = country)) +
  geom_line(size = 1) +
  ggtitle("Top 6 Countries with Highest Average Deaths ") +
  labs(x = "year", y = "Deaths per 100,000")
```

From the line chart, we can see the top 6 countries with highest average yearly deaths are all from Africa.Although there is an increase of average deaths from 1990-2005 for four countries, a rapid decrease can be seen from 2010 onwards. In general, the situation is getting better compared to the start year 1990.

## Data Visualization
Firstly, a choropleth map is perfect to show a clear regional pattern in data, which in line with malaria case. So I create a choropleth map to have a full picture of malaria deaths situation over the world (using plotly)

```{r}
malaria_death_processed_hover <- 
  malaria_death_processed %>%
  mutate(hover = paste0(country, "\ndeaths: ", deaths))

fig_choropleth = plot_geo(malaria_death_processed_hover,
                     frame = ~year) %>%
  add_trace(locations = ~code,
            z = ~deaths,
            zmin = 0,
            zmax = max(malaria_death_processed$deaths),
            colors = "Reds",
            text = ~hover,
            hoverinfo = 'text') %>%
  layout(#geo = list(scope = 'africa'),
         title = 'Malaria Deaths around the World\n1990-2016') %>%
  config(displayModeBar = FALSE)

# First visualization
fig_choropleth
```

From the choropleth map, we can see that Africa is the most affected region by malaria. But the death rate has gradually dropped over time, especially in recent years since the dark red color changes to a very light one. 

Next, Let's move on to malaria_death_age.csv dataset to analyse deaths across different age groups

```{r message=FALSE, warning = FALSE}
# Data cleaning
malaria_death_age$...1 <- NULL 
malaria_death_age_processed <- malaria_death_age %>%
  setNames(c("country","code","year","age_groups","deaths"))

fig_age <- malaria_death_age_processed %>%
  group_by(year, age_groups) %>%
  summarise(avg_yearly_death = mean(deaths))

fig_age$age_groups <- factor(fig_age$age_groups,levels = c("Under 5", "5-14", "15-49", "50-69", "70 or older")) 

# A stacked bar chart is used here
fig_bar <- ggplot(fig_age, aes(year, avg_yearly_death)) +
  geom_col(aes(fill = age_groups)) +
  scale_fill_brewer(direction = -1) +
  labs (x = "year", y = "Deaths", title = "Malaria Death by Age Group\n1990-2016") 

ggplotly(fig_bar) %>%
  config(displayModeBar = FALSE)
```

From the stacked bar chart, we can see that children under 5 is the majority affected age group who die from malaria while 70 or older elderly people are rarely affected by malaria. The older you are, the less likely you are to die from malaria. 

Finally, I'd like to use a scatter plot to see the relationship between incidence and deaths. They should have positively correlated normally. Since Africa is the most affected region based the previous analysis result, I only filter out countries in Africa to draw the plot.

```{r}
# pre-processing: add continent/region column to the dataset
malaria_inc_processed <- malaria_inc %>%
  setNames(c("country","code","year","incidence")) %>%
  filter(!is.na(code)) %>%
  mutate(region = countrycode(code,'iso3c','region')) %>%
  filter(!is.na(region))
```

```{r}
# join the malaria_inc and malaria_death together using leftjoin
malaria_inc_death_combined <- malaria_inc_processed %>%
  left_join(malaria_death_processed, by = c('country','code', 'year'))

# Filter out Africa data
Africa_df <- malaria_inc_death_combined %>%
  filter(region == 'Middle East & North Africa' | region == 'Sub-Saharan Africa')

fig_scatter <- ggplot(Africa_df, aes(incidence, deaths, color = year)) +
  geom_point() +
  #geom_smooth() +
  facet_wrap(~year)

fig_scatter
```

From the scatter plot, we can confirm the positive relationship between malaria incidence and deaths. However, when we look into year 2000 and 2005, we can see they show a exponential pattern while year 2010 and 2015 have more linear pattern. Exponential pattern is not a good sign because it represents the number of deaths will increase faster but decrease slower compared to the incidence.

